### 2024-04-22 paper/paper.qmd
finished making graphs and updated cleaning script

---
title: "My title"
subtitle: "My subtitle if needed"
author: "Faiza Imam"
thanks: "Code and data are available at: []."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(knitr)
library(tidyverse)
library(usethis)
library(kableExtra)
library (arrow)
library(opendatatoronto)
library(janitor)

library(here)
home_data <- readr::read_csv(here::here("data/analysis_data/clean_dataNA.csv"))

```

# Introduction

You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.

The remainder of this paper is structured as follows. @sec-data....

# Data {#sec-data}

Some of our data is of penguins (@fig-bills), from @palmerpenguins.
```{r}
#| echo: false
#| label: tbl-summ_table
#| tbl-cap: "Summary of clean data"
#| message: false

Sum_data <- home_data %>%
  group_by(Year_of_death, Cause_of_death, Age_group, Gender) %>%
  summarize(total_deaths = sum(death_toll))

# Write the summarized data to a CSV file
write_csv(x = Sum_data, file = "Sum_data.csv")

head(Sum_data, 7) %>%
  kable(col.names = c("Year","Cause of Death", "Age Group", "Gender", "Death Toll"), booktabs = TRUE)

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

Talk way more about it.

```{r}
#| echo: false
#| label: tbl-Gender
#| tbl-cap: "Homelessness gender total death tolls"

#Gender Death Tolls
GendervDeath <- home_data %>%
  group_by(Gender) %>%
  summarize(total_deaths = sum(death_toll))

 write_csv(
  x = GendervDeath ,
  file = "GendervDeath.csv"
)

  head (GendervDeath, 7) |>
  kable(
    col.names = c("Gender", "Death Tolls"),
  booktabs = TRUE
  )

```

```{r}
#| echo: false
#| label: tbl-Agegroup
#| tbl-cap: "Age Groups and their total death tolls"

#Age group death count
AgevDeath <- home_data %>%
  group_by(Age_group) %>%
  summarize(total_deaths = sum(death_toll))

 write_csv(
  x = AgevDeath ,
  file = "AgevDeath.csv"
)

  head (AgevDeath, 7) |>
  kable(
    col.names = c("Age Group", "Death Tolls"),
  booktabs = TRUE
  )
```

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| warning: false
#| label: fig-Yearvcuase
#| fig-cap: Cause of death by year

############
#Yearly death causes
YearvCause <- home_data %>%
  group_by(Year_of_death,Cause_of_death) %>%
  summarize(total_deaths = sum(death_toll))

 write_csv(
  x = YearvCause,
  file = "YearvCause.csv"
)

 YearvCause %>%
  ggplot(aes(x = Year_of_death , y = total_deaths, fill = factor(Cause_of_death))) +
  geom_col(position = "dodge", alpha = 0.8) +  # Use geom_col() for bar charts
  labs(title = "Total Cause of death by year", x = "Year", y = "Total Deaths", fill = "Cause of Death") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
#| label: fig-AgevCause
#| fig-cap: Total cause of death by age group
#| echo: false
#| message: false

# Define the order of age groups
age_order <- c("<20", "20-39", "40-59", "60+")

# Use factor to set the order of levels for Age_group
home_data$Age_group <- factor(home_data$Age_group, levels = age_order)

# Group by Cause_of_death and Age_group, maintaining the age order
AgevCause <- home_data %>%
  group_by(Cause_of_death, Age_group) %>%
  summarize(total_deaths = sum(death_toll))

 write_csv(
  x = AgevCause,
  file = "AgevCause.csv"
)

 AgevCause %>%
  ggplot(aes(x = Age_group , y = total_deaths, fill = factor(Cause_of_death))) +
  geom_col(position = "dodge", alpha = 0.8) + 
  labs(title = "Total deaths by Age Group", x = "Age Group", y = "Total Deaths", fill = "Cause of deaths") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
#| label: fig-Gendercause
#| fig-cap: "Total cause of deaths grouped by gender"
#| echo: false
#| message: false

#Gender and cuases of death within each group
GendervCause <- home_data %>%
  group_by(Cause_of_death, Gender) %>%
  summarize(total_deaths = sum(death_toll))

 write_csv(
  x = GendervCause ,
  file = "GendervCause.csv"
)

GendervCause %>%
  ggplot(aes(x = Cause_of_death, y = total_deaths, fill = factor(Gender))) +
  geom_col(position = "dodge", alpha = 0.8) + 
  labs(title = "Cause of Death by Gender", x = "Cause of Death", y = "Total Deaths", fill = "Gender") +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

# References

### 2024-04-22 scripts/02-data_cleaning.R

@@ -56,9 +56,11 @@ clean_omitNA <-
   select(
     Year_of_death,
     Gender,
-    Age,
-    death_toll
+    Age_group,
+    death_toll,
   ) 
+
+
 
 
 #Following few lines of code are from  https://tellingstorieswithdata.com/02-drinking_from_a_fire_hose.html

### 2024-04-22 scripts/02-data_cleaning.R
Cleaning updated (again)

@@ -58,6 +58,7 @@ clean_omitNA <-
     Gender,
     Age_group,
     death_toll,
+    Cause_of_death
   ) 
 
 

### 2024-04-24 paper/datasheet_template.qmd
Updated and Uploaded datasheet + Refs

---
title: "Datasheet for Unveiling Vulnerabilities"
subtitle: "An Analysis of Causes of Death Across Age Groups and Genders Within Toronto's Homelessness Population"
author: "Faiza Imam"
thanks: "Code and data are available at: https://github.com/fazim07/deaths_gender"
date: today
date-format: long
abstract: "This datasheet describes the motivations and composition of homesless death by cause data set obtained from OpenData Toronto. It also describes collection process, cleaning process, uses, distribution, and maintenance of the data sets."
format: pdf
number-sections: true
bibliography: references.bib
---

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
     -The dataset "Deaths of People Experiencing Homelessness" was created to track the deaths of people experiencing homelessness more accurately and understand their causes. The specific task was to improve data collection and analysis in order to gain insights into the mortality rates and causes of death within the homeless population in Toronto.

2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    -The dataset was created by Toronto Public Health (TPH) as part of their initiative to track and analyze homeless deaths. TPH oversees the data collection, analysis, and reporting, with support from the Shelter, Support and Housing Administration (SSHA) and various health and social service agencies that assist the homeless population. The Office of the Chief Coroner of Ontario (OCCO) also verifies some of the data.

3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    -The funding for the creation of this dataset is not specified in the provided information.

4. *Any other comments?*
    -There are no additional comments provided at this time.

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- The instances in the dataset represent deaths of people experiencing homelessness in Toronto. Each instance likely includes information such as the cause of death, age group, year of death, gender.
2. *How many instances are there in total (of each type, if appropriate)?*
	- The total number of instances in the dataset is not specified in the information provided. The file contains 253 entries in total.

3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- Is the dataset a sample or representative of a larger set? The dataset represents reported deaths of people experiencing homelessness in Toronto and is not necessarily a random sample. It reflects data collected and reported by Toronto Public Health (TPH), the Shelter, Support and Housing Administration (SSHA), and other community partners, and is subject to limitations such as unknown causes of death, missing Indigenous status information, and unreported deaths.

4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	-  Each instance likely consists of structured data fields such as cause of death, age group, year of death, gender, and possibly additional information related to the circumstances of the death.

5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- The main target or focus of the dataset is the cause of death among people experiencing homelessness, with additional variables providing context such as age group, gender, and year of death.
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- Yes, there are limitations mentioned regarding missing information such as unknown causes of death, missing Indigenous status, missed or unreported age groups, races, ethnicities, and unreported deaths.
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- The information from Opendata provided does not mention explicit relationships between individual instances.

8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- No specific data splits or recommendations are mentioned in the provided information.
Are there any errors, noise, or redundancies in the dataset?

9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- The dataset is subject to limitations such as unknown causes of death, missing Indigenous status information, and unreported deaths, which could be considered sources of noise or potential errors.

10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	-The dataset appears to be self-contained, based on the provided information on Opendata Toronto data portal.

11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- The dataset likely contains sensitive information regarding individuals' deaths, but specific details about confidentiality are not provided but can assumed to be confidential.

12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	  - The dataset contains information related to deaths, which may be sensitive but is not explicitly described as offensive or threatening.

13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	-The dataset includes sub-populations based on age groups and genders. The gender variable covers three entries; male, female and unknown. The *unknown* entries are not classfied nor described. The age group variable covers four demographs, <20, 20-39, 40-59, 60+.

14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- The dataset likely does not contain personally identifiable information such as names or specific identifiers, there is an *id* number which is associated with every entry.
Does the dataset contain sensitive data? 
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- The dataset contains sensitive data related to deaths, causes of death, and demographic information, but specifics about sensitive categories like race, sexual orientation, etc., are not provided. The cause of death variable covers 9 common cuases of death within Toronto's homeless population.
16. *Any other comments?*
	- No comment

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- The data was acquired by TPH collecting reports of deaths among the homeless population in Toronto. Therefore there maybe some gaps due to individual being miscounted or uncounted for.

2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- No addtional information regarding data collection is provided in the description.

3. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- Toronto Public Health, no other informaiton is provided.

4. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- Data collection was a binannual collection effort.

5. *Any other comments?*
	- No Comments

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- No informaiton regarding data cleaning is provided

2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- Raw data is saved here;https://github.com/fazim07/deaths_gender/tree/master/data/raw_data

3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- Statistical programming language R was used to code and clean the data; https://github.com/fazim07/deaths_gender/blob/master/scripts/02-data_cleaning.R

4. *Any other comments?*
	- No comment

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- No tasks have been provided to indicate as such. Possible tasks may be municipal reports regarding this data.

2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- No repository is provided to suggest as such.

3. *What (other) tasks could the dataset be used for?*
	- Dataset could be used to analyze further trends across ethnicity, race and sociao-economic status. Additionally, the data can provide insight into the policy and vital information for the Government when making laws.

4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- All the data from set is confidential and contains death counts and sensitive data on cause of death that may be trigger to future users.

5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- No, unless the intention is malicious and the intend to spread misinformation.
6. *Any other comments?*
	- No comment

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- While being hosted on Opendata portal, its free access to the public hence a third party may have access to it.

2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- The data set comes in a CSV, JSON or Excel file no DOIs.

3. *When will the dataset be distributed?*
	- The latest update or distribution was on Apr 17, 2024

4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- Dataset is licensed under Open Government Licence – Toronto - https://open.toronto.ca/open-data-license/

5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- No third party IP based or other restrictions on the data association and instances

6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- Cannot access who these individuals are that have passed.

7. *Any other comments?*
	- No comment

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- Opendata Toronto portal will host and support the dataset while Opendata Toronto updates the data set.
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- Can be contacted via email from Opendata Toronto

3. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- Yes, this data updates in a end of year manner adding new collected satistics of the homelessness populations. The data collection will be collected by TPH and maintained and stored in Opendata Toronto.

4. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- The data refelects on the homeless peoples deaths, however, other than the cause, age and gender there is no other additional personal information provided hence no need for retention.

5. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- Older versions of the data are supported and maintained as the study is continues to be collected and updated by the end of the year by the TPH.

6. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- TBD
7. *Any other comments?*
	- No Comments

\newpage

### 2024-04-24 paper/paper.qmd
saving

@@
 ---
-title: "My title"
-subtitle: "My subtitle if needed"
+title: "Unveiling Vulnerabilities"
+subtitle: "An Analysis of Causes of Death Across Age Groups and Genders Within Toronto's Homelessness Population"
 author: "Faiza Imam"
-thanks: "Code and data are available at: []."
+thanks: "Code and data are available at: https://github.com/fazim07/deaths_gender"
 date: today
@@
 library(knitr)
-library(tidyverse)
 library(usethis)
 library(kableExtra)
-library (arrow)
 library(opendatatoronto)
 library(janitor)
+library(here)
+library(dplyr)
+library(ggplot2)
 
-library(here)
 home_data <- readr::read_csv(here::here("data/analysis_data/clean_dataNA.csv"))
@@
 # Introduction
+Homelessness is typically thought of as an individual who is completely unhoused or resides in public spaces.
+Four categories are commonly used to describe homelessness to various degrees.
+Chronic homelessness is a pattern in which an individual has had or is facing homelessness in a continuous pattern.
+Often, these victims are older and struggle with complex health issues, i.e., diabetes, mental illness, or addiction.
+Episodic homelessness refers to an individual who has experienced 3-4 periods of homelessness within a year [@statCan].
+Typically seen in younger cohorts, these individuals often fall prey to addiction or have mental health issues.
+This specific category is critical as episodic homelessness can transform into chronic if left unprovoked [@4homes].
+Transitional homelessness refers to individuals who become homeless due to a major life event that affects their financial, familial, or social situation.
+Considered the most common form of homelessness, particularly since the pandemic, it is typically found in younger working individuals and can lead them to seek shelters for temporary housing [@4homes].
+Lastly, hidden homelessness refers to individuals who are 'couch-surfing' with friends or family.
+This type is typically undocumented and unreported, hence not much data is collected from it.
+Individuals facing this particular type have no guarantee that they may find housing and are not registered under any shelters [@4homes].
 
-You can and should cross-reference sections and sub-sections. We use @citeR and @rohan.
+Some of the key factors that result in homelessness are financial challenges, results from abuse or intimate partner violence, health issues which cannot be sustained within a household, and finally addiction or substance abuse (@statCan).
+Although many underlying factors contribute to an individual becoming or facing homelessness, it’s important to keep an open and empathetic  mind when dealing with the data collected to understand these trends.
+Research has shown that women make up 27.3% of the homeless population.
+Several factors contribute to the higher likelihood of women experiencing homelessness compared to men.
+These factors include unstable employment, which raises the risk of income fluctuations or job loss, as well as the added responsibility of care giving, such as caring for children and dependents, intimate partner violence and other forms of abuse (@Homehubs).
 
-The remainder of this paper is structured as follows. @sec-data....
+This paper aims to identify the correlation between gender and age within the homeless population in Toronto from the year 2017 - 2023, specifically examining how age and gender intersect in terms of homelessness experiences and demographic characteristics.
+In examining the data set concerning causes of death, including homelessness as a potential factor, I hypothesize that there is a notable relationship between female homelessness and mortality rates across different age groups.
+Specifically, the anticipation that younger age groups (<20 and 20-39) among homeless females may exhibit higher mortality rates compared to older age groups (40-59 and 60+), potentially due to factors such as exposure to harsh living conditions, limited access to healthcare, and increased vulnerability to external threats.
+
+The paper follows a structured approach comprising an [Introduction], [Data], [Results], and [Discussion] sections.
+The [Introduction] sets the stage by providing essential background information on homelessness in Toronto and outlining the research objectives.
+Moving into the [Data] section, the methodology for data collection, including gender, age, and homelessness status assessment.
+The [Results] section presents findings on the correlation between gender and age within Toronto's homeless population, using visual aids to depict trends and significant patterns.
+Finally, the [Discussion] section interprets these results in the context of the research questions, compares them with existing literature, addresses methodological limitations and biases, and suggests potential areas for future research to further expand understanding in this domain.
+
+The graphs and tables in this paper were made with [@citeR] using R studio.
+The creation of these graphs and tables was made with `ggplot` [@ggplot], `tiddyverse` [@rTidy], The analysis and cleaning of the data were conducted with and `dyplr` [@dply], `janitor` [@janitor] and `readr` [@readr] packages.
 
@@
 
-Some of our data is of penguins (@fig-bills), from @palmerpenguins.
+The data was collected from Open Data Toronto, specifically from the data set called "Deaths of People Experiencing Homelessness," which is a subset known as "Homeless deaths by cause." Toronto Public Health (TPH) conducted this data collection initiative to track the deaths of people experiencing homelessness more accurately and understand their causes.
+TPH oversees the data collection, analysis, and reporting, while the Shelter, Support and Housing Administration (SSHA) and various health and social service agencies that support the homeless population share information with TPH.
+The Office of the Chief Coroner of Ontario (OCCO) also verifies some of the data.
+In this context, homelessness is defined as "the situation of an individual or family without stable, permanent, appropriate housing, or the immediate prospect, means, and ability to acquire it [@Odata].
+
+## Data Source
+
+The original data set contained six variables, for this paper, I have selected five of the six.
+The first variable, *id* contains an id number which is associated with the following variables, *cause of death*, *age group*, *year of death*, and *gender*.
+There are a total of 253 entries.
+The *cause of death* variable is listened from 12 categories.
+These categories include homicide, accident, cardiovascular disease, cancer, COVID-19, drug toxicity, pneumonia, suicide, others and unknown.
+There are no notices nor indications on what the category other and unknown classifies as.
+The *age group* variable is a set of four age ranges, <20 years old, 20-39 years of age, 40-59 years of age and 60+ years of age @tbl-Agegroup.
+*Year of death*, covers the years from 2017 to 2023.
+Lastly, the gender variable houses, three categories, male, female, and unknown @tbl-Gender.
+Once again, the *unknown* has not been classified.
+
+## Variables of Intrest & Cleaning
+
+To clean the data, five of the six variables were chosen as the *id* variable was not required for this paper.
+From there the *age group* category had unknown variables.
+To avoid any issues with the gender death analysis, the unknown variables were omitted from the clean data set.
+This brings the clean data set to 233 entries as opposed to the original 253, @tbl-summtable shows a snap shot of the clean data sets.
+The decision to omit unknown values in the age group category is particularly important when analyzing how gender correlates with different causes of death across specific age groups.
+Including unknown values could introduce uncertainty and skew the findings, making it challenging to draw accurate conclusions about any relationships or trends observed between gender, age, and causes of death among the homeless population.
+
+Therefore, by focusing on known age group data, the analysis can provide more insightful and reliable insights into how gender may influence mortality rates or patterns of causes of death within different age brackets among the homeless population in Toronto.
+
 ```{r}
 #| echo: false
-#| label: tbl-summ_table
+#| label: tbl-summtable
 #| tbl-cap: "Summary of clean data"
@@
 ```
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
-
-Talk way more about it.
 
@@
 
+age_order <- c("<20", "20-39", "40-59", "60+")
+
+# Use factor to set the order of levels for Age_group
+home_data$Age_group <- factor(home_data$Age_group, levels = age_order)
+
+# Group by Cause_of_death and Age_group, maintaining the age order
+
 #Age group death count
@@
 
-# Results
-
-Our results are summarized in @tbl-modelresults.
-
 ```{r}
@@
 #| label: fig-Yearvcuase
-#| fig-cap: Cause of death by year
+#| fig-cap: "Cause of death by year trends form 2017 - 2023"
 
-############
 #Yearly death causes
@@
   geom_col(position = "dodge", alpha = 0.8) +  # Use geom_col() for bar charts
-  labs(title = "Total Cause of death by year", x = "Year", y = "Total Deaths", fill = "Cause of Death") +
+  labs(title = "Yearly cause of death trends", x = "Cause of Death", y = "Total Deaths", fill = "Cause of death") +
   theme_minimal() + 
@@
 ```{r}
-#| label: fig-Gendercause
-#| fig-cap: "Total cause of deaths grouped by gender"
+#| label: agedeathsbyyear
+#| fig-cap: "Total deaths by age group from 2017-2023"
+#| fig-subcap: ["Total Deaths by Age in 2017", "Total Deaths by Age in 2018", "Total Deaths by Age in 2019", "Total Deaths by Age in 2020", "Total Deaths by Age in 2021","Total Deaths by Age in 2022","Total Deaths by Age in 2023"]
+#| layout-ncol: 2
 #| echo: false
+#| warning: false
 #| message: false
 
-#Gender and cuases of death within each group
-GendervCause <- home_data %>%
-  group_by(Cause_of_death, Gender) %>%
+#2017 dATA
+home2017 <- home_data %>%
+  filter(Year_of_death == 2017) %>%
+  count(Cause_of_death, Age_group) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2017, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Age_group)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+#########2018
+home2018 <- home_data %>%
+  filter(Year_of_death == 2018) %>%
+  count(Cause_of_death, Age_group) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2018, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Age_group)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+#######2019
+
+home2019 <- home_data %>%
+  filter(Year_of_death == 2019) %>%
+  count(Cause_of_death, Age_group) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2019, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Age_group)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2020
+home2020 <- home_data %>%
+  filter(Year_of_death == 2020) %>%
+  count(Cause_of_death, Age_group) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2020, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Age_group)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2021
+home2021 <- home_data %>%
+  filter(Year_of_death == 2021) %>%
+  count(Cause_of_death, Age_group) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2021, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Age_group)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2022
+home2022 <- home_data %>%
+  filter(Year_of_death == 2022) %>%
+  count(Cause_of_death, Age_group) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2022, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Age_group)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2023
+home2023 <- home_data %>%
+  filter(Year_of_death == 2023) %>%
+  count(Cause_of_death, Age_group) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2023, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Age_group)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+```
+
+```{r}
+#| label: genderyear
+#| fig-cap: "Total deaths by age group from 2017-2023"
+#| fig-subcap: ["Total Deaths by Age in 2017", "Total Deaths by Age in 2018", "Total Deaths by Age in 2019", "Total Deaths by Age in 2020", "Total Deaths by Age in 2021","Total Deaths by Age in 2022","Total Deaths by Age in 2023"]
+#| layout-ncol: 2
+#| echo: false
+#| warning: false
+#| message: false
+
+##### 2017
+home2017 <- home_data %>%
+  filter(Year_of_death == 2017) %>%
+  count(Cause_of_death, Gender) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2017, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Gender)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2018
+home2018 <- home_data %>%
+  filter(Year_of_death == 2018) %>%
+  count(Cause_of_death, Gender) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2018, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Gender)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2019
+home2019 <- home_data %>%
+  filter(Year_of_death == 2019) %>%
+  count(Cause_of_death, Gender) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2019, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Gender)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2019
+home2019 <- home_data %>%
+  filter(Year_of_death == 2019) %>%
+  count(Cause_of_death, Gender) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2019, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Gender)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2020
+home2020 <- home_data %>%
+  filter(Year_of_death == 2020) %>%
+  count(Cause_of_death, Gender) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2020, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Gender)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2021
+home2021 <- home_data %>%
+  filter(Year_of_death == 2021) %>%
+  count(Cause_of_death, Gender) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2021, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Gender)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2022
+home2022 <- home_data %>%
+  filter(Year_of_death == 2022) %>%
+  count(Cause_of_death, Gender) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2022, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Gender)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+
+##### 2023
+home2023 <- home_data %>%
+  filter(Year_of_death == 2023) %>%
+  count(Cause_of_death, Gender) %>%
+  top_n(10, n)  # Filter for the top 10 causes of death
+
+# Plot the top 10 causes of death
+ggplot(home2023, aes(x = fct_reorder(Cause_of_death, n), y = n, fill = Gender)) +
+  geom_bar(stat = "identity") +
+  theme_minimal() +
+  theme(legend.position = "bottom", axis.text.x = element_text(angle = 45, hjust = 1)) +
+  labs(x = "Age Group", y = "Death Count", fill = "Cause of Death")
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-AgeGencause
+#| fig-cap: Looking at the death toll trends of age groups and gender
+
+AgeGencause <- home_data %>%
+  group_by(Age_group, Gender) %>%
   summarize(total_deaths = sum(death_toll))
@@
  write_csv(
-  x = GendervCause ,
-  file = "GendervCause.csv"
+  x = AgeGencause,
+  file = "AgeGencause.csv"
 )
 
-GendervCause %>%
-  ggplot(aes(x = Cause_of_death, y = total_deaths, fill = factor(Gender))) +
-  geom_col(position = "dodge", alpha = 0.8) + 
-  labs(title = "Cause of Death by Gender", x = "Cause of Death", y = "Total Deaths", fill = "Gender") +
+AgeGencause %>%
+  ggplot(aes(x = Age_group, y = total_deaths, fill = factor(Gender))) +
+  geom_col(alpha = 0.8) +  # Use geom_col() for bar charts
+  labs(title = "Deaths by Age group to Gender", x = "Age Group", y = "Total Deaths", fill = "Gender") +
+  scale_fill_manual(values = c("maroon","lightblue", "purple")) +  # Specify colors for each Gender level
   theme_minimal() + 
@@
 ```
+# Results
 
@@
 
-## First discussion point {#sec-first-point}
+## Yearly Death Analysis
 
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
+## Causes of Death by Age Group
 
-## Second discussion point
+## Causes of Death by Gender
 
-## Third discussion point
+## Weaknesses & Next Steps
+When considering weaknesses and potential biases in research on the correlation between gender and age within the homeless population in Toronto, several aspects should be addressed to understand what factors may have skewed the data projections during this paper's analysis.
+There was limited data available regarding age and gender breakdowns, and even with the available data, there were a few entries that were unknown for the age category.
+I opted to exclude those values to focus on the existing and available values within the dataset.
+This decision could have skewed the data and constrained the accuracy of the findings.
+As mentioned previously, there is an underlying stigma surrounding homelessness, which alone can influence individuals who are reluctant to disclose information, particularly those facing hidden homelessness.
+Lastly, within the gender values, there are three variables: male, female, and unknown.
+This unknown entry possibly refers to those individuals who have an LGBTQ+ identity and can overlook important nuances in the experiences of homeless individuals.
+Addressing these weaknesses and biases through rigorous study design, comprehensive data collection methods, ethical considerations, and thoughtful analysis can enhance the reliability and validity of research findings in this area."
 
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
+Future studies in this area should aim to investigate the intersectionality of gender, race, ethnicity LGBTQ+ identity, disability status and mental illness to understand the long-term effects of homelessness within these demographics.
+Findings from this research can help understand how policymakers and organizations can help alleviate the growing numbers of homelessness victims and pave the way for research methodologies and intervention strategies aimed at addressing homelessness in a broader sense.
+Another future study can be aimed at exploring the long-term impacts of homelessness on an individual's physical and mental health, as well as their ability to reintegrate into stable housing and employment situations whether they face chronic or episodic homelessness.
+Understanding the nuanced challenges faced by homeless individuals, including the effects on physical and mental health, reintegration into stable housing and employment, and the cyclical nature of homelessness, is essential for creating effective and sustainable solutions.
+By prioritizing research and intervention strategies that consider the multifaceted aspects of homelessness, we can work towards a more inclusive and supportive approach that fosters long-term stability and well-being for vulnerable populations.
 

### 2024-04-24 paper/paper.qmd
saving readme and paper pdf + qmd

@@
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper aims to explore the relationship between male homelessness and mortality rates across age groups, focusing on causes of death related to homelessness in Toronto from the years 2017- 2023.
+The hypothesis suggests that younger homeless males (<20 and 20-39) may have higher mortality rates due to harsh living conditions, limited healthcare access, and increased vulnerability.
+The results indicate that while male homeless mortality rates are generally higher than reported female death tolls, the most affected age demographic is the 40-59 age range rather than the hypothesized age group."
 format: pdf
@@
 This paper aims to identify the correlation between gender and age within the homeless population in Toronto from the year 2017 - 2023, specifically examining how age and gender intersect in terms of homelessness experiences and demographic characteristics.
-In examining the data set concerning causes of death, including homelessness as a potential factor, I hypothesize that there is a notable relationship between female homelessness and mortality rates across different age groups.
-Specifically, the anticipation that younger age groups (<20 and 20-39) among homeless females may exhibit higher mortality rates compared to older age groups (40-59 and 60+), potentially due to factors such as exposure to harsh living conditions, limited access to healthcare, and increased vulnerability to external threats.
+In examining the data set concerning causes of death, including homelessness as a potential factor, I hypothesize that there is a notable relationship between male homelessness and mortality rates across different age groups.
+Specifically, the anticipation that younger age groups (<20 and 20-39) among homeless males may exhibit higher mortality rates compared to older age groups (40-59 and 60+), potentially due to factors such as exposure to harsh living conditions, limited access to healthcare, and increased vulnerability to external threats.
 
@@
 There are no notices nor indications on what the category other and unknown classifies as.
-The *age group* variable is a set of four age ranges, <20 years old, 20-39 years of age, 40-59 years of age and 60+ years of age @tbl-Agegroup.
 *Year of death*, covers the years from 2017 to 2023.
@@
 Once again, the *unknown* has not been classified.
+We can observe by the table that Male homeless victims have highest death tolls sitting at 749, while female victims come in second with 210, followed by unknown indivudals who are accounted for as 3 The *age group* variable is a set of four age ranges, <20 years old, 20-39 years of age, 40-59 years of age and 60+ years of age @tbl-Agegroup.
+We can see that there are 9 deaths reported for the age gorup under 20, followed by the 20-39 age range, then the 60+ age range and finally, the 40-59 age range having the highest death tolls among each demograph.
 
@@
   summarize(total_deaths = sum(death_toll))
+AgevCause
 
@@
 ```{r}
+#| echo: false
+#| warning: false
+#| message: false
+#| label: fig-AgeGencause
+#| fig-cap: Looking at the death toll trends of age groups and gender
+
+AgeGencause <- home_data %>%
+  group_by(Age_group, Gender) %>%
+  summarize(total_deaths = sum(death_toll))
+
+ write_csv(
+  x = AgeGencause,
+  file = "AgeGencause.csv"
+)
+
+AgeGencause %>%
+  ggplot(aes(x = Age_group, y = total_deaths, fill = factor(Gender))) +
+  geom_col(alpha = 0.8) +  # Use geom_col() for bar charts
+  labs(title = "Deaths by Age group to Gender", x = "Age Group", y = "Total Deaths", fill = "Gender") +
+  scale_fill_manual(values = c("maroon","lightblue", "purple")) +  # Specify colors for each Gender level
+  theme_minimal() + 
+  theme(axis.text.x = element_text(angle = 45, hjust = 1))
+
+```
+
+```{r}
 #| label: genderyear
-#| fig-cap: "Total deaths by age group from 2017-2023"
-#| fig-subcap: ["Total Deaths by Age in 2017", "Total Deaths by Age in 2018", "Total Deaths by Age in 2019", "Total Deaths by Age in 2020", "Total Deaths by Age in 2021","Total Deaths by Age in 2022","Total Deaths by Age in 2023"]
+#| fig-cap: "Total Deaths by Gender from 2017-2023"
+#| fig-subcap: ["Total Deaths by Gender in 2017", "Total Deaths by Gender in 2018", "Total Deaths by Gender in 2019", "Total Deaths by Gender in 2020", "Total Deaths by Gender in 2021","Total Deaths by Gender in 2022","Total Deaths by Gender in 2023"]
 #| layout-ncol: 2
@@
 
-```{r}
-#| echo: false
-#| warning: false
-#| message: false
-#| label: fig-AgeGencause
-#| fig-cap: Looking at the death toll trends of age groups and gender
+# Results
 
-AgeGencause <- home_data %>%
-  group_by(Age_group, Gender) %>%
-  summarize(total_deaths = sum(death_toll))
+## Yearly Death Analysis
 
- write_csv(
-  x = AgeGencause,
-  file = "AgeGencause.csv"
-)
+## Age Groups and Cause of Death
 
-AgeGencause %>%
-  ggplot(aes(x = Age_group, y = total_deaths, fill = factor(Gender))) +
-  geom_col(alpha = 0.8) +  # Use geom_col() for bar charts
-  labs(title = "Deaths by Age group to Gender", x = "Age Group", y = "Total Deaths", fill = "Gender") +
-  scale_fill_manual(values = c("maroon","lightblue", "purple")) +  # Specify colors for each Gender level
-  theme_minimal() + 
-  theme(axis.text.x = element_text(angle = 45, hjust = 1))
-
-```
-# Results
+## Gender and Cause of Death
 
@@
 
-## Yearly Death Analysis
+## Male to Female Cause of Death Trends
 
-## Causes of Death by Age Group
+## Age Demograph Cause of Death Trends
 
-## Causes of Death by Gender
+## Male Age Specific Death Trends
 

### 2024-04-24 paper/paper.qmd
updating paper and pdf will filled sections. Updated LLM

@@
 These factors include unstable employment, which raises the risk of income fluctuations or job loss, as well as the added responsibility of care giving, such as caring for children and dependents, intimate partner violence and other forms of abuse (@Homehubs).
+However, there is no study explaining the reason to the high volume of male victims in homelessness populations.
 
 This paper aims to identify the correlation between gender and age within the homeless population in Toronto from the year 2017 - 2023, specifically examining how age and gender intersect in terms of homelessness experiences and demographic characteristics.
-In examining the data set concerning causes of death, including homelessness as a potential factor, I hypothesize that there is a notable relationship between male homelessness and mortality rates across different age groups.
+In examining the data set concerning causes of death, including homelessness as a potential factor, the hypothesis suggested is that there is a notable relationship between male homelessness and mortality rates across different age groups.
 Specifically, the anticipation that younger age groups (<20 and 20-39) among homeless males may exhibit higher mortality rates compared to older age groups (40-59 and 60+), potentially due to factors such as exposure to harsh living conditions, limited access to healthcare, and increased vulnerability to external threats.
 
-The paper follows a structured approach comprising an [Introduction], [Data], [Results], and [Discussion] sections.
 The [Introduction] sets the stage by providing essential background information on homelessness in Toronto and outlining the research objectives.
@@
 
-The original data set contained six variables, for this paper, I have selected five of the six.
+The original data set contained six variables, for this paper, five of the six variables were selected.
 The first variable, *id* contains an id number which is associated with the following variables, *cause of death*, *age group*, *year of death*, and *gender*.
@@
 Once again, the *unknown* has not been classified.
-We can observe by the table that Male homeless victims have highest death tolls sitting at 749, while female victims come in second with 210, followed by unknown indivudals who are accounted for as 3 The *age group* variable is a set of four age ranges, <20 years old, 20-39 years of age, 40-59 years of age and 60+ years of age @tbl-Agegroup.
+We can observe by the table that Male homeless victims have highest death tolls sitting at 749, while female victims come in second with 210, followed by unknown individuals who are accounted for as 3 The *age group* variable is a set of four age ranges, <20 years old, 20-39 years of age, 40-59 years of age and 60+ years of age @tbl-Agegroup.
 We can see that there are 9 deaths reported for the age gorup under 20, followed by the 20-39 age range, then the 60+ age range and finally, the 40-59 age range having the highest death tolls among each demograph.
@@
   summarize(total_deaths = sum(death_toll))
-AgevCause
 
@@
 ```{r}
-#| label: agedeathsbyyear
+#| label: fig-agedeathsbyyear
 #| fig-cap: "Total deaths by age group from 2017-2023"
@@
 ```{r}
-#| label: genderyear
+#| label: fig-genderyear
 #| fig-cap: "Total Deaths by Gender from 2017-2023"
@@
 
+Analyzing the data from the figure @fig-Yearvcuase, we can observe trends in the yearly death toll among the homeless population from 2017 to 2023.
+When sorting by the cause of death, we find that drug toxicity stands out as the most prevalent cause of death during this period.
+From 2017 until the beginning of 2023, drug toxicity consistently accounted for the highest number of reported deaths among the homeless.
+In 2021, we see a notable peak in reported drug-related deaths, with 123 cases recorded.
+This spike underscores the severity of substance abuse issues within the homeless community during that year.
+Following drug toxicity as a leading cause of death, we find that unknown causes emerge as the second most common reason for mortality.
+The data shows that in 2023, there was a significant increase in reported deaths with unknown causes, reaching a peak of 51 cases.
+This rise in unknown causes of death raises questions about the accuracy and completeness of reporting or may indicate underlying health issues or circumstances that require further investigation.
+Overall, these insights highlight critical areas for intervention and support within homeless populations, particularly addressing substance abuse and improving data collection and reporting mechanisms to better understand and address mortality trends.
+
 ## Age Groups and Cause of Death
+
+When analyzing the overall death trends over the years based on age groups @fig-AgevCause, a distinct pattern emerges, particularly regarding the age demographics of 20-39 and 40-59, which exhibit the highest death tolls.
+The primary cause attributed to these deaths is drug toxicity, indicating a significant challenge within these age brackets.
+Specifically, the 20-39 age group had a death toll of 137, while the 40-59 age group recorded a death toll of 143, highlighting the critical impact of substance abuse and related health risks in these younger to middle-aged homeless individuals.
+Moreover, projecting this trend forward, we observe that the second-highest cause of death among the 60+ age group is classified as unknown (@fig-AgevCause).
+This finding underscores the complexity of factors contributing to mortality in older homeless individuals, pointing to potential gaps in understanding and addressing the specific health issues or circumstances leading to these unknown causes of death.
+
+When examining the figures provided in @fig-agedeathbyyear, a clear picture emerges regarding the prevalent causes of death among various age demographics over the years.
+One notable trend is the consistent reporting of the highest death tolls among the 40-59 age demographic across different years.
+In particular, the data reveals that in 2019 @agedeathbyyear (c ), this age group experienced the highest number of deaths in seven out of the nine cause-of-death categories considered.
+This finding underscores the significant impact of mortality within the 40-59 age range and suggests a need for targeted interventions and health strategies to address the specific challenges faced by individuals in this demographic.
+Further analysis of these trends can offer valuable insights into the dynamics of mortality rates among different age groups, guiding efforts to improve healthcare access, address underlying health issues, and enhance support systems for vulnerable populations, including those experiencing homelessness.
+
+These insights prompt further exploration into the underlying factors driving mortality rates across different age groups within the homeless population, emphasizing the need for targeted interventions and support strategies tailored to address the diverse and evolving health challenges faced by individuals experiencing homelessness.
 
@@
 
+In the gender distribution depicted in @fig-AgeGencause, male victims are predominant across all age demographics.
+Specifically, the 20-39 age group recorded 170 reported male deaths compared to 80 reported female deaths, resulting in a male-to-female ratio of approximately 4:5 within this age demographic.
+Moving to the 40-59 age group, the data shows 331 male deaths and 88 female deaths, translating to a male-to-female death ratio of approximately 2:1.
+This indicates a higher proportion of male deaths relative to females within this age range.
+The most significant skew in gender distribution is observed in the 60+ age demographic, where there were 244 male deaths compared to 37 female deaths.
+This disparity yields a male-to-female death ratio of approximately 7:1, representing the highest skew among all the age demographics analyzed.
+
+When analyzing the figures provided in @fig-genderyear from a gender perspective, a prominent pattern emerges highlighting the dominance of male deaths across various age demographics over the years.
+One striking trend is the consistent reporting of the highest death tolls among the 40-59 age demographic, primarily driven by male fatalities, which persist across different years.
+This emphasizes the substantial impact of mortality among males within the 40-59 age range, indicating the urgent need for targeted interventions and health strategies.
+
+These results emphasize the urgent need for targeted interventions and health strategies, particularly in addressing substance abuse issues and understanding the gender-specific dynamics of mortality within the homeless population.
+
 # Discussion
 
-## Male to Female Cause of Death Trends
+This paper aims to delve into the relationship between gender and age within Toronto's homeless population with a specific focus on mortality rates and causes of death.
+The central hypothesis posits that male homelessness is significantly correlated with higher mortality rates across different age groups.
+The anticipation of elevated mortality rates among younger homeless males, especially those below 20 and between 20 to 39, stems from the recognition of harsh living conditions, limited healthcare accessibility, and increased vulnerability to external risks among this demographic.
 
-## Age Demograph Cause of Death Trends
+## Male-Female Cause of Death Trends
 
-## Male Age Specific Death Trends
+The analysis of gender-specific cause-of-death trends among Toronto's homeless population from 2017 to 2023 reveals notable patterns.
+Across various age groups, males consistently exhibit higher mortality rates compared to females.
+This trend is particularly pronounced in the 20-39 and 40-59 age brackets, where male deaths significantly outnumber female deaths.
+The data underscores the disproportionate impact of homelessness on male individuals within these age demographics.
+One possible explanation for this disparity could be the differential exposure to risk factors and vulnerabilities among male and female homeless individuals.
+Factors such as substance abuse, violence, and lack of access to adequate health care may contribute to higher mortality rates among males.
+Additionally, societal norms and systemic challenges may disproportionately affect males in terms of seeking and receiving support services.
+The reasoning behind such a high volume of men being reported as homeless is unclear and should be studied moving forward to understand how homelessness affects this demographic so harshly.
+Based on the results and findings, it can be assumed that most of these reported deaths among homeless individuals are due to episodic and chronic homelessness.
+Although that particular information is not provided, it should be a consideration for understanding how often the population flows fluctuate.
+
+The findings highlight the importance of addressing gender-specific health disparities within homeless populations.
+Policymakers should make note and create targeted interventions focusing on addressing substance abuse, improving access to healthcare, and providing comprehensive support services tailored to the needs of male individuals.
+This in turn can play a crucial role in mitigating mortality rates and improving overall well-being to mitigate the growing homeless population.
+
+## Male Age-Specific Cause of Death Trends
+
+The analysis of male age-specific cause of death trends within Toronto's homeless population from 2017 to 2023 reveals nuanced patterns that warrant discussion.
+Across different age brackets, certain causes of death emerge as prominent contributors to mortality rates among male homeless individuals.
+In all reported cases, drug toxicity takes the top spot, with an unknown cause of death ranking second.
+Although the exact nature of this cause is unknown, further studies should be conducted to dissect and identify the underlying issue.
+
+In the younger age groups, particularly those below 20 and between 20 to 39, drug toxicity stands out as a leading cause of death.
+This finding aligns with broader trends indicating the prevalence of substance abuse issues among homeless individuals, especially among males in their early adulthood.
+The impact of harsh living conditions, limited healthcare accessibility, and heightened vulnerability to external risks likely exacerbates the risk of drug-related fatalities in these demographics.
+With easy access to unregulated drugs and substances, it's apparent that these deaths result from overdoses.
+Moving to the 40-59 age group, while drug toxicity remains a significant cause of death, other factors such as chronic health conditions and external injuries also contribute substantially to mortality rates among males.
+The interplay of long-term health challenges, environmental hazards, and socioeconomic factors underscores the complexity of mortality risks faced by middle-aged male homeless individuals.
+In the 60+ age demographic, unknown causes of death emerge as a notable concern among male homeless individuals.
+This finding raises questions about the accuracy and completeness of reporting, as well as the potential presence of undiagnosed or untreated health conditions among older homeless males.
+The lack of visibility into the specific causes of death in this age group highlights a gap in understanding the health needs and challenges of elderly homeless individuals.
+
+The main hypothesis of this study suggests a significant connection between male homelessness and mortality rates, focusing particularly on the 20-39 age group.
+Initial analysis affirmed a notable impact on male homeless individuals compared to females.
+However, the hypothesis fails to account for variations within gender and age subsets.
+For instance, younger homeless males aged 20-39 may face increased mortality risks due to substance abuse, whereas older homeless males may encounter different mortality patterns related to chronic health conditions and age-related vulnerabilities.
+To enhance future research, a more nuanced approach is required, considering factors like substance use patterns, healthcare accessibility, social support networks, and environmental exposures.
+Integrating these factors will lead to a more comprehensive understanding of mortality risks among homeless populations and inform targeted interventions.
 

### 2024-04-24 paper/paper.qmd
Updated paper and qmd

@@
 number-sections: true
+toc: true
 bibliography: references.bib
@@
 
-Some of the key factors that result in homelessness are financial challenges, results from abuse or intimate partner violence, health issues which cannot be sustained within a household, and finally addiction or substance abuse (@statCan).
+Some of the key factors that result in homelessness are financial challenges, results from abuse or intimate partner violence, health issues which cannot be sustained within a household, and finally addiction or substance abuse [@statCan].
 Although many underlying factors contribute to an individual becoming or facing homelessness, it’s important to keep an open and empathetic  mind when dealing with the data collected to understand these trends.
@@
 Several factors contribute to the higher likelihood of women experiencing homelessness compared to men.
-These factors include unstable employment, which raises the risk of income fluctuations or job loss, as well as the added responsibility of care giving, such as caring for children and dependents, intimate partner violence and other forms of abuse (@Homehubs).
+These factors include unstable employment, which raises the risk of income fluctuations or job loss, as well as the added responsibility of care giving, such as caring for children and dependents, intimate partner violence and other forms of abuse [@Homehubs].
 However, there is no study explaining the reason to the high volume of male victims in homelessness populations.
@@
 
-When examining the figures provided in @fig-agedeathbyyear, a clear picture emerges regarding the prevalent causes of death among various age demographics over the years.
+When examining the figures provided in @fig-agedeathsbyyear, a clear picture emerges regarding the prevalent causes of death among various age demographics over the years.
 One notable trend is the consistent reporting of the highest death tolls among the 40-59 age demographic across different years.
-In particular, the data reveals that in 2019 @agedeathbyyear (c ), this age group experienced the highest number of deaths in seven out of the nine cause-of-death categories considered.
+In particular, the data reveals that in 2019 @fig-agedeathsbyyear (c), this age group experienced the highest number of deaths in seven out of the nine cause-of-death categories considered.
 This finding underscores the significant impact of mortality within the 40-59 age range and suggests a need for targeted interventions and health strategies to address the specific challenges faced by individuals in this demographic.
@@
 
-## Weaknesses & Next Steps
+# Weaknesses & Next Steps
 When considering weaknesses and potential biases in research on the correlation between gender and age within the homeless population in Toronto, several aspects should be addressed to understand what factors may have skewed the data projections during this paper's analysis.

### 2024-04-25 scripts/00-simulate_data.R
Fixed typos in scripts

@@ -6,7 +6,7 @@
 # License: MIT
 # Pre-requisites: None
 
-#### Workspace setup ####
+#### Workspace setup #### 
 library(tidyverse)
 library(tibble)
 

### 2024-04-25 scripts/01-download_data.R

@@ -10,7 +10,7 @@ library(opendatatoronto)
 library(tidyverse)
 library(here)
 library (arrow)
-library(dplyr)
+library(dplyr) 
 #### Download data ####
 
 #The following lines of code is from https://open.toronto.ca/dataset/deaths-of-people-experiencing-homelessness/

### 2024-04-25 scripts/03-test_data.R

@@ -8,7 +8,7 @@
 
 
 #### Workspace setup ####
-library(tidyverse)
+library(tidyverse) 
 
 ## Test if there are 7 years
 clean_omitNA$Year_of_death |> min(na.rm = TRUE) >= 7

### 2024-04-25 paper/datasheet_template.qmd
Reviewed and fixed typos in datasheet

@@
 date-format: long
-abstract: "This datasheet describes the motivations and composition of homesless death by cause data set obtained from OpenData Toronto.
-It also describes collection process, cleaning process, uses, distribution, and maintenance of the data sets."
+abstract: "This datasheet describes the motivations and composition of homesless death by cause dataset obtained from OpenData Toronto.
+It also describes collection process, cleaning process, uses, distribution, and maintenance of the datasets."
 format: pdf
@@
 2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
-	- The data set comes in a CSV, JSON or Excel file no DOIs.
+	- The dataset comes in a CSV, JSON or Excel file no DOIs.
 
@@
 1. *Who will be supporting/hosting/maintaining the dataset?*
-	- Opendata Toronto portal will host and support the dataset while Opendata Toronto updates the data set.
+	- Opendata Toronto portal will host and support the dataset while Opendata Toronto updates the dataset.
 2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*

### 2024-04-25 paper/paper.qmd
Final round of edit for paper

@@
 Analyzing the data from the figure @fig-Yearvcuase, we can observe trends in the yearly death toll among the homeless population from 2017 to 2023.
-When sorting by the cause of death, we find that drug toxicity stands out as the most prevalent cause of death during this period.
+When sorting by the cause of death, we find that drug toxicity stands out as the most prevalent cause of death throughout the years.
 From 2017 until the beginning of 2023, drug toxicity consistently accounted for the highest number of reported deaths among the homeless.
@@
 
-When examining the figures provided in @fig-agedeathsbyyear, a clear picture emerges regarding the prevalent causes of death among various age demographics over the years.
+When examining the figures provided in @fig-agedeathsbyyear, a pattern emerges regarding the prevalent causes of death among various age demographics over the years.
 One notable trend is the consistent reporting of the highest death tolls among the 40-59 age demographic across different years.
@@
 This finding underscores the significant impact of mortality within the 40-59 age range and suggests a need for targeted interventions and health strategies to address the specific challenges faced by individuals in this demographic.
-Further analysis of these trends can offer valuable insights into the dynamics of mortality rates among different age groups, guiding efforts to improve healthcare access, address underlying health issues, and enhance support systems for vulnerable populations, including those experiencing homelessness.
-
-These insights prompt further exploration into the underlying factors driving mortality rates across different age groups within the homeless population, emphasizing the need for targeted interventions and support strategies tailored to address the diverse and evolving health challenges faced by individuals experiencing homelessness.
+Further analysis of these trends can offer valuable insights into the dynamics of mortality rates among different age groups, guiding efforts to improve healthcare access for the homeless population in Toronto.
 
@@
 
-In the gender distribution depicted in @fig-AgeGencause, male victims are predominant across all age demographics.
-Specifically, the 20-39 age group recorded 170 reported male deaths compared to 80 reported female deaths, resulting in a male-to-female ratio of approximately 4:5 within this age demographic.
-Moving to the 40-59 age group, the data shows 331 male deaths and 88 female deaths, translating to a male-to-female death ratio of approximately 2:1.
-This indicates a higher proportion of male deaths relative to females within this age range.
+In the gender distribution depicted in @fig-AgeGencause, male victims are dominant across all age demographics.
+Specifically, the 20-39 age group which recorded 170 reported male deaths compared to 80 reported female deaths, resulting in a male-to-female ratio of approximately 4:5 within this age demographic.
+Moving to the 40-59 age group, the data shows 331 male reported deaths and 88 reported female deaths, translating to a male-to-female death ratio of approximately 2:1.
+This indicates a higher proportion of male death reports relative to females within this age range.
 The most significant skew in gender distribution is observed in the 60+ age demographic, where there were 244 male deaths compared to 37 female deaths.
-This disparity yields a male-to-female death ratio of approximately 7:1, representing the highest skew among all the age demographics analyzed.
+This skew yields a male-to-female death ratio of approximately 7:1, representing the highest skew among all the age demographics analyzed within this dataset.
 
 When analyzing the figures provided in @fig-genderyear from a gender perspective, a prominent pattern emerges highlighting the dominance of male deaths across various age demographics over the years.
-One striking trend is the consistent reporting of the highest death tolls among the 40-59 age demographic, primarily driven by male fatalities, which persist across different years.
+One striking trend is the consistent reporting of the highest death tolls among the 40-59 age demographic, primarily driven by male fatalities, which persist across different years, the highest being 2017.
 This emphasizes the substantial impact of mortality among males within the 40-59 age range, indicating the urgent need for targeted interventions and health strategies.
-
-These results emphasize the urgent need for targeted interventions and health strategies, particularly in addressing substance abuse issues and understanding the gender-specific dynamics of mortality within the homeless population.
 
@@
 The data underscores the disproportionate impact of homelessness on male individuals within these age demographics.
-One possible explanation for this disparity could be the differential exposure to risk factors and vulnerabilities among male and female homeless individuals.
+One possible explanation for this disparity could be the exposure to risk factors and vulnerabilities among male and female homeless individuals.
 Factors such as substance abuse, violence, and lack of access to adequate health care may contribute to higher mortality rates among males.
 Additionally, societal norms and systemic challenges may disproportionately affect males in terms of seeking and receiving support services.
-The reasoning behind such a high volume of men being reported as homeless is unclear and should be studied moving forward to understand how homelessness affects this demographic so harshly.
+The reasoning behind such a high volume of men being reported as homeless is unclear and should be studied moving forward to understand how homelessness affects this specific demographic so harshly.
 Based on the results and findings, it can be assumed that most of these reported deaths among homeless individuals are due to episodic and chronic homelessness.
@@
 
-The analysis of male age-specific cause of death trends within Toronto's homeless population from 2017 to 2023 reveals nuanced patterns that warrant discussion.
+The analysis of male age-specific cause of death trends within Toronto's homeless population from 2017 to 2023 reveals nuanced patterns that raises discussion.
 Across different age brackets, certain causes of death emerge as prominent contributors to mortality rates among male homeless individuals.
 In all reported cases, drug toxicity takes the top spot, with an unknown cause of death ranking second.
-Although the exact nature of this cause is unknown, further studies should be conducted to dissect and identify the underlying issue.
+Although the exact nature of this cause is unknown, further studies should be conducted to dissect and identify the underlying issues.
 
@@
 With easy access to unregulated drugs and substances, it's apparent that these deaths result from overdoses.
-Moving to the 40-59 age group, while drug toxicity remains a significant cause of death, other factors such as chronic health conditions and external injuries also contribute substantially to mortality rates among males.
-The interplay of long-term health challenges, environmental hazards, and socioeconomic factors underscores the complexity of mortality risks faced by middle-aged male homeless individuals.
+Moving to the 40-59 age group, while drug toxicity remains a significant cause of death, other factors such as chronic health conditions and external injuries also contribute substantially to mortality rates among males but are not noted.
+The emphasis of long-term health challenges, environmental hazards, and socioeconomic factors underscores the complexity of mortality risks faced by middle-aged male homeless individuals.
 In the 60+ age demographic, unknown causes of death emerge as a notable concern among male homeless individuals.
-This finding raises questions about the accuracy and completeness of reporting, as well as the potential presence of undiagnosed or untreated health conditions among older homeless males.
-The lack of visibility into the specific causes of death in this age group highlights a gap in understanding the health needs and challenges of elderly homeless individuals.
+This finding raises questions about the accuracy and completeness of dataset, as well as the potential presence of undiagnosed or untreated health conditions among older homeless males that contribute to the high deathrate compared to their female counterparts.
+The lack of visibility into the specific causes of death in this age group highlights a gap in understanding the health needs and challenges of elderly homeless individuals in a broader perspective.
 
@@
 Initial analysis affirmed a notable impact on male homeless individuals compared to females.
-However, the hypothesis fails to account for variations within gender and age subsets.
+However, the hypothesis fails at the notion of younf adult amles (20-39) being of higher death tolls.
 For instance, younger homeless males aged 20-39 may face increased mortality risks due to substance abuse, whereas older homeless males may encounter different mortality patterns related to chronic health conditions and age-related vulnerabilities.
